---
title: "IL-10 suppression of inflammatory chromatin in primary human monocytes"
subtitle: "Downstream ATAC-seq analysis of Mishra et al. GSE261244"
author: "Swathi Dhanasekaran"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 140
)
set.seed(20260509)
options(stringsAsFactors = FALSE)
```

```{r libraries}
needed <- c(
  "limma", "GenomicRanges", "rtracklayer",
  "BSgenome.Hsapiens.UCSC.hg38", "JASPAR2024", "TFBSTools", "motifmatchr",
  "AnnotationDbi", "org.Hs.eg.db",
  "ggplot2", "dplyr", "tidyr", "readr", "tibble", "purrr",
  "pheatmap", "ggrepel", "patchwork", "R.utils"
)
missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install missing packages before rendering: ", paste(missing, collapse = ", "))
}

library(limma)
library(GenomicRanges)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg38)
library(JASPAR2024)
library(TFBSTools)
library(motifmatchr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(pheatmap)
library(ggrepel)
library(patchwork)
library(R.utils)
```

```{bash manifest}
echo "Input manifest"
wc -l mishra_il10_atac_takehome/GSE261244_Merged_ATAC_peaks.bed
wc -l mishra_il10_atac_takehome/GSE261244_ATAC_normalized_counts.txt
shasum -a 256 mishra_il10_atac_takehome/GSE261244_Merged_ATAC_peaks.bed
shasum -a 256 mishra_il10_atac_takehome/GSE261244_ATAC_normalized_counts.txt
```

```{r paths}
data_dir <- "mishra_il10_atac_takehome"
counts_file <- file.path(data_dir, "GSE261244_ATAC_normalized_counts.txt")
bed_file <- file.path(data_dir, "GSE261244_Merged_ATAC_peaks.bed")
ref_dir <- "references"
dir.create(ref_dir, showWarnings = FALSE)

gencode_gtf_gz <- file.path(ref_dir, "gencode.v45.annotation.gtf.gz")
gencode_gtf_url <- "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_45/gencode.v45.annotation.gtf.gz"
hallmark_gmt <- file.path(ref_dir, "h.all.v2024.1.Hs.symbols.gmt")
hallmark_url <- "https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2024.1.Hs/h.all.v2024.1.Hs.symbols.gmt"

if (!file.exists(gencode_gtf_gz)) {
  download.file(gencode_gtf_url, gencode_gtf_gz, mode = "wb", quiet = TRUE)
}
if (!file.exists(hallmark_gmt)) {
  download.file(hallmark_url, hallmark_gmt, mode = "wb", quiet = TRUE)
}
```

# Overview {-}

This analysis uses the paper-released, library-size-normalized, log-scale ATAC
peak matrix from primary human CD14+ monocytes. I use limma-trend as the
differential accessibility engine because it models continuous log-scale
accessibility across all six conditions while retaining donor as a blocking
factor. I do not use DESeq2 or edgeR raw-count GLMs because rounding normalized
log values back to integers would create artificial count data.

The biological question is not simply "which peaks are significant?" It is
whether IL-10 closes the inflammatory chromatin response wholesale, or whether it
preferentially suppresses an IRF/interferon-linked component while much of the
NF-kB/AP-1 inflammatory architecture remains accessible. That is the nuance the
paper argues for, and it is the nuance this reduced ATAC-seq bundle can test.

# Peak set and sample matrix characterization

```{r load-inputs}
counts <- read.delim(counts_file, row.names = 1, check.names = FALSE)
bed <- read.delim(
  bed_file,
  header = FALSE,
  col.names = c("chrom", "start", "end")
)

# README: BED is 0-based, half-open. Therefore width is end - start, and the
# count matrix row name should be chr:start-end without any +1 conversion.
bed <- bed %>%
  mutate(
    peak_id = paste0(chrom, ":", start, "-", end),
    width = end - start
  )

stopifnot(nrow(counts) == 73341)
stopifnot(nrow(bed) == 73341)
stopifnot(identical(rownames(counts), bed$peak_id))
stopifnot(ncol(counts) == 24)
stopifnot(all(bed$width > 0))

sample_info <- tibble(sample = colnames(counts)) %>%
  mutate(
    condition = sub("_D[0-9]+$", "", sample),
    donor = sub("^.*_(D[0-9]+)$", "\\1", sample),
    condition = factor(
      condition,
      levels = c("Rest", "IL10", "LPS", "IL10.plus.LPS", "TNF", "IL10.plus.TNF")
    ),
    donor = factor(donor, levels = paste0("D", 1:4))
  )

stopifnot(n_distinct(sample_info$condition) == 6)
stopifnot(n_distinct(sample_info$donor) == 4)
stopifnot(all(table(sample_info$condition) == 4))
stopifnot(all(table(sample_info$donor) == 6))

matrix_summary <- tibble(
  peaks = nrow(counts),
  samples = ncol(counts),
  min_value = min(as.matrix(counts)),
  median_value = median(as.matrix(counts)),
  mean_value = mean(as.matrix(counts)),
  max_value = max(as.matrix(counts)),
  missing_values = sum(is.na(counts))
)
matrix_summary
```

```{r peak-characterization}
chrom_summary <- bed %>%
  count(chrom, name = "n_peaks") %>%
  mutate(chrom = factor(chrom, levels = c(paste0("chr", 1:22), "chrX", "chrY"))) %>%
  arrange(chrom)

width_summary <- bed %>%
  summarise(
    min_width = min(width),
    median_width = median(width),
    mean_width = mean(width),
    max_width = max(width)
  )
width_summary

ggplot(chrom_summary, aes(chrom, n_peaks)) +
  geom_col(fill = "#356d8c") +
  labs(x = NULL, y = "Merged peaks", title = "Peak distribution across hg38 chromosomes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(bed, aes(width)) +
  geom_histogram(bins = 60, fill = "#6f8f3f", color = "white") +
  labs(x = "Peak width (bp; BED end - start)", y = "Peaks", title = "Merged peak widths") +
  theme_bw()
```

```{r accessibility-distribution}
long_counts <- counts %>%
  rownames_to_column("peak_id") %>%
  pivot_longer(-peak_id, names_to = "sample", values_to = "accessibility") %>%
  left_join(sample_info, by = "sample")

ggplot(long_counts, aes(accessibility, color = condition)) +
  geom_density(linewidth = 0.7) +
  labs(
    x = "Log-scale normalized accessibility",
    y = "Density",
    title = "Sample accessibility distributions are broadly comparable"
  ) +
  theme_bw()
```

The input checks support using this bundle as a coherent downstream ATAC-seq
object: `r nrow(counts)` peaks, `r ncol(counts)` samples, no missing values, and
an exact ordered 1:1 match between the BED intervals and matrix rows. The BED
coordinates are used as standard 0-based, half-open intervals for ID matching;
only when constructing `GRanges` for Bioconductor annotation do I convert starts
to 1-based coordinates.

The merged peak set has a median width of `r median(bed$width)` bp and spans
autosomes plus chrX/chrY, with no chrM peaks in the consensus set. That peak-size
range is compatible with a merged ATAC regulatory-element set rather than broad
histone domains. The accessibility distributions are smooth and broadly aligned
across conditions, which argues against gross sample-level failure or global
chromatin collapse. These checks cannot prove the original libraries passed
fragment-level ENCODE metrics, but they do confirm that the released matrix is
internally consistent and biologically suitable for donor-paired downstream
analysis.

## MCQ 1: peak and matrix characterization

The released matrix contains 73,341 merged hg38 ATAC peaks across 24 samples,
with smooth log-scale accessibility distributions and an exact ordered match to
the 3-column BED file. Which interpretation best captures what this first
characterization establishes biologically?

A. The dataset is ready for downstream chromatin-state comparison because the
peak universe and sample matrix are internally coherent, the distributions do
not suggest global sample failure, and the peak widths are consistent with
regulatory ATAC elements.

B. The dataset proves that every upstream ENCODE ATAC-seq QC metric passed,
because chrM peaks are absent and the normalized matrix has no missing values.

C. The most important biological feature of the matrix is the exact number of
peaks; once the dimensions match the README, downstream interpretation is
independent of coordinate convention and sample distribution.

D. The broad range of log accessibility values means the matrix should be
back-transformed and rounded so that raw-count models can be used for all
subsequent contrasts.

**Answer: A.** The section establishes that the released object is internally
valid and interpretable as a regulatory ATAC peak matrix. It does not recreate
fragment-level QC, and it does not justify raw-count models on normalized
log-scale values.

# Sample QC and replicate structure

```{r qc-plots}
sample_cor <- cor(as.matrix(counts), method = "pearson")
pheatmap(
  sample_cor,
  annotation_col = as.data.frame(sample_info[, c("sample", "condition", "donor")]) %>%
    tibble::column_to_rownames("sample"),
  annotation_row = as.data.frame(sample_info[, c("sample", "condition", "donor")]) %>%
    tibble::column_to_rownames("sample"),
  main = "Sample-sample Pearson correlations",
  fontsize = 8
)

pca <- prcomp(t(as.matrix(counts)), center = TRUE, scale. = FALSE)
pca_df <- as_tibble(pca$x[, 1:4], rownames = "sample") %>%
  left_join(sample_info, by = "sample")
var_explained <- (pca$sdev^2 / sum(pca$sdev^2))[1:4]
qc_summary <- tibble(
  min_pairwise_correlation = min(sample_cor[upper.tri(sample_cor)]),
  median_pairwise_correlation = median(sample_cor[upper.tri(sample_cor)]),
  pc1_variance = var_explained[1],
  pc2_variance = var_explained[2],
  pc3_variance = var_explained[3],
  pc4_variance = var_explained[4]
)
qc_summary

ggplot(pca_df, aes(PC1, PC2, color = condition, shape = donor, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 2.6, max.overlaps = 24) +
  labs(
    x = paste0("PC1 (", scales::percent(var_explained[1], accuracy = 0.1), ")"),
    y = paste0("PC2 (", scales::percent(var_explained[2], accuracy = 0.1), ")"),
    title = "PCA of normalized ATAC accessibility"
  ) +
  theme_bw()

ggplot(pca_df, aes(PC3, PC4, color = condition, shape = donor, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 2.6, max.overlaps = 24) +
  labs(
    x = paste0("PC3 (", scales::percent(var_explained[3], accuracy = 0.1), ")"),
    y = paste0("PC4 (", scales::percent(var_explained[4], accuracy = 0.1), ")"),
    title = "Later PCs check donor and treatment structure"
  ) +
  theme_bw()
```

```{r donor-pair-qc}
condition_means <- long_counts %>%
  group_by(condition, donor) %>%
  summarise(median_accessibility = median(accessibility), .groups = "drop")

ggplot(condition_means, aes(condition, median_accessibility, group = donor, color = donor)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  labs(
    x = NULL,
    y = "Median log accessibility",
    title = "Global accessibility summaries retain the paired donor design"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

The QC plots ask whether the biological comparison is credible before any
formal contrast is fit. The samples show high pairwise correlation
(minimum `r round(qc_summary$min_pairwise_correlation, 3)`, median
`r round(qc_summary$median_pairwise_correlation, 3)`) with no obvious isolated
sample. PCA separates samples by treatment structure rather than producing an
IL-10-only failure cluster. That is important biologically: IL-10 is not acting
like a nonspecific toxicity or global chromatin-closing condition in this matrix.

The donor-wise median accessibility plot shows that the same four donors remain
trackable across all six conditions. This supports the paired design and
protects the analysis from mistaking donor baseline differences for stimulus or
IL-10 effects. The upstream ENCODE-style fragment metrics are not present in the
bundle, so the appropriate downstream conclusion is narrower but still useful:
the released normalized matrix behaves like a coherent, paired ATAC-seq dataset
for contrast analysis, with no sample that should be excluded at this stage.

## MCQ 2: sample QC and ENCODE context

After distribution, correlation, PCA and donor-pair checks, which conclusion
best describes the sample-level structure of this ATAC-seq matrix?

A. IL-10-containing samples behave as global accessibility outliers, supporting
the idea that IL-10 broadly shuts down chromatin rather than selectively
modulating stimulus-responsive regions.

B. The samples are sufficiently coherent for paired downstream analysis:
replicates correlate well, no sample is an obvious outlier, and donor identity
remains visible across conditions without overwhelming the treatment structure.

C. Donor pairing can be ignored because PCA already shows treatment separation,
so the donor term would only remove biological signal.

D. The matrix-level QC proves that TSS enrichment, FRiP and insert-size
periodicity met ENCODE thresholds, so no further caveat is needed.

**Answer: B.** The sample structure supports the intended paired contrast
design. The result also argues against a global IL-10 accessibility artifact,
which is important for interpreting later suppression signatures as selective
regulatory modulation rather than sample failure.

# Contrast design and differential accessibility

```{r limma-design}
design <- model.matrix(~ 0 + condition + donor, sample_info)
colnames(design) <- sub("^condition", "", colnames(design))

contrast_matrix <- makeContrasts(
  LPS_vs_Rest = LPS - Rest,
  TNF_vs_Rest = TNF - Rest,
  IL10_vs_Rest = IL10 - Rest,
  IL10LPS_vs_LPS = IL10.plus.LPS - LPS,
  IL10TNF_vs_TNF = IL10.plus.TNF - TNF,
  levels = design
)

contrast_plan <- tibble(
  contrast = c(
    "LPS_vs_Rest",
    "TNF_vs_Rest",
    "IL10_vs_Rest",
    "IL10LPS_vs_LPS",
    "IL10TNF_vs_TNF"
  ),
  biological_question = c(
    "Which regulatory elements open or close during TLR4-driven inflammatory stimulation?",
    "Which regulatory elements respond to TNF receptor signaling in the same donor-matched cells?",
    "What chromatin changes are induced by IL-10 alone before inflammatory challenge?",
    "Which LPS-responsive elements are suppressed when IL-10 is present?",
    "Does the IL-10 suppression direction generalize in the TNF context?"
  )
)
contrast_plan

fit <- lmFit(as.matrix(counts), design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE)

contrast_results <- map(
  colnames(contrast_matrix),
  \(coef_name) {
    topTable(fit2, coef = coef_name, number = Inf, sort.by = "none") %>%
      rownames_to_column("peak_id") %>%
      mutate(contrast = coef_name)
  }
) %>%
  bind_rows()

da_summary <- contrast_results %>%
  mutate(direction = case_when(
    adj.P.Val < 0.05 & logFC > 0.25 ~ "higher",
    adj.P.Val < 0.05 & logFC < -0.25 ~ "lower",
    TRUE ~ "not_FDR_effect"
  )) %>%
  count(contrast, direction) %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0)
da_summary
```

```{r volcanoes}
volcano_df <- contrast_results %>%
  mutate(
    label = case_when(
      adj.P.Val < 0.05 & abs(logFC) > 1 ~ "FDR < 0.05, |logFC| > 1",
      adj.P.Val < 0.05 & abs(logFC) > 0.25 ~ "FDR < 0.05, |logFC| > 0.25",
      TRUE ~ "Other"
    ),
    neg_log10_fdr = -log10(pmax(adj.P.Val, .Machine$double.xmin))
  )

ggplot(volcano_df, aes(logFC, neg_log10_fdr, color = label)) +
  geom_point(alpha = 0.25, size = 0.6) +
  facet_wrap(~ contrast, scales = "free_y") +
  scale_color_manual(values = c(
    "FDR < 0.05, |logFC| > 1" = "#9f2d20",
    "FDR < 0.05, |logFC| > 0.25" = "#d18f2f",
    "Other" = "grey75"
  )) +
  labs(x = "limma log fold-change", y = "-log10 FDR", color = NULL, title = "Donor-adjusted accessibility contrasts") +
  theme_bw()
```

The five contrasts separate three biological questions. `LPS vs Rest` and
`TNF vs Rest` define the stimulus-responsive chromatin programs. `IL10 vs Rest`
tests whether IL-10 alone preconditions the resting monocyte accessibility
landscape. The two suppression contrasts ask whether adding IL-10 reduces
accessibility after LPS or TNF challenge.

I call directionally interpretable peaks at FDR < 0.05 and absolute logFC >
0.25. The FDR threshold controls the large number of peak-wise tests, while the
effect-size threshold avoids treating extremely small but statistically stable
changes as biologically central. At that threshold, both LPS and TNF produce
large accessibility programs, which is expected for primary monocytes exposed to
strong innate immune stimuli. IL-10 reduces more peaks than it increases in the
stimulated settings (`r da_summary$lower[da_summary$contrast == "IL10LPS_vs_LPS"]`
lower versus `r da_summary$higher[da_summary$contrast == "IL10LPS_vs_LPS"]`
higher for LPS; `r da_summary$lower[da_summary$contrast == "IL10TNF_vs_TNF"]`
lower versus `r da_summary$higher[da_summary$contrast == "IL10TNF_vs_TNF"]`
higher for TNF), consistent with a suppressive chromatin effect.

The paired donor term is essential because each donor contributes all six
conditions. This is normally what I would want in a small primary-cell ATAC-seq
study: biological donor variability is real, but the paired design lets the
model ask whether the same donor shifts accessibility after stimulation or
IL-10. The next step is therefore not to compare all significant peaks
indiscriminately, but to ask which stimulus-induced elements are reversed by
IL-10 and whether the LPS and TNF arms show coherent directionality.

## MCQ 3: contrast design

The donor-adjusted contrasts show strong LPS and TNF chromatin responses and
more IL-10-associated decreases than increases in the stimulated settings. Which
interpretation best uses this contrast structure?

A. The key IL-10 biology should be inferred from `IL10 vs Rest`, because
anti-inflammatory memory must be visible before stimulation if it is real.

B. The correct suppression question is stimulus-conditioned: define LPS- and
TNF-responsive accessibility changes first, then ask whether IL-10 reverses
those changes within the same donor-blocked design.

C. Since both LPS and TNF generate many significant peaks, the exact identities
of suppressed peaks are not biologically meaningful and downstream motif
analysis should use all significant peaks together.

D. The donor term should be excluded because primary donor variability is a
clinical nuisance rather than part of the biological design.

**Answer: B.** The contrast design keeps the biology conditional: IL-10
suppression is interpreted against the stimulus program it modifies, not as an
unpaired or stimulus-agnostic shift in accessibility.

# IL-10 suppression signature

```{r suppression-signature}
lps_res <- contrast_results %>% filter(contrast == "LPS_vs_Rest")
tnf_res <- contrast_results %>% filter(contrast == "TNF_vs_Rest")
lps_sup_res <- contrast_results %>% filter(contrast == "IL10LPS_vs_LPS")
tnf_sup_res <- contrast_results %>% filter(contrast == "IL10TNF_vs_TNF")

signature_tbl <- bed %>%
  select(peak_id, chrom, start, end, width) %>%
  left_join(lps_res %>% select(peak_id, lps_logFC = logFC, lps_fdr = adj.P.Val), by = "peak_id") %>%
  left_join(tnf_res %>% select(peak_id, tnf_logFC = logFC, tnf_fdr = adj.P.Val), by = "peak_id") %>%
  left_join(lps_sup_res %>% select(peak_id, il10_lps_logFC = logFC, il10_lps_fdr = adj.P.Val), by = "peak_id") %>%
  left_join(tnf_sup_res %>% select(peak_id, il10_tnf_logFC = logFC, il10_tnf_fdr = adj.P.Val), by = "peak_id") %>%
  mutate(
    lps_induced = lps_fdr < 0.05 & lps_logFC > 1,
    tnf_induced = tnf_fdr < 0.05 & tnf_logFC > 1,
    il10_lps_suppressed = lps_induced & il10_lps_fdr < 0.05 & il10_lps_logFC < -0.75,
    il10_lps_broad_suppressed = lps_fdr < 0.05 & lps_logFC > 0.25 & il10_lps_fdr < 0.05 & il10_lps_logFC < -0.25,
    il10_tnf_same_direction = il10_tnf_logFC < 0
  )

suppression_summary <- signature_tbl %>%
  summarise(
    lps_induced = sum(lps_induced),
    tnf_induced = sum(tnf_induced),
    strong_lps_il10_suppressed = sum(il10_lps_suppressed),
    broad_lps_il10_suppressed = sum(il10_lps_broad_suppressed),
    fraction_strong_signature_same_direction_in_tnf =
      mean(il10_tnf_same_direction[il10_lps_suppressed])
  )
suppression_summary

foreground_peaks <- signature_tbl %>% filter(il10_lps_suppressed)
broad_foreground_peaks <- signature_tbl %>% filter(il10_lps_broad_suppressed)
```

```{r suppression-plots}
ggplot(signature_tbl, aes(lps_logFC, il10_lps_logFC)) +
  geom_point(alpha = 0.16, size = 0.5, color = "grey45") +
  geom_point(
    data = signature_tbl %>% filter(il10_lps_suppressed),
    aes(lps_logFC, il10_lps_logFC),
    alpha = 0.45, size = 0.7, color = "#9f2d20"
  ) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(
    x = "LPS vs Rest logFC",
    y = "IL10 + LPS vs LPS logFC",
    title = "IL-10 suppression is concentrated among LPS-induced peaks"
  ) +
  theme_bw()

ggplot(signature_tbl %>% filter(il10_lps_suppressed), aes(il10_lps_logFC, il10_tnf_logFC)) +
  geom_point(alpha = 0.3, size = 0.7, color = "#356d8c") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(
    x = "IL10 + LPS vs LPS logFC",
    y = "IL10 + TNF vs TNF logFC",
    title = "TNF contrast is used as a sign-consistency check"
  ) +
  theme_bw()
```

```{r signature-heatmap}
top_signature <- foreground_peaks %>%
  arrange(il10_lps_fdr, il10_lps_logFC) %>%
  slice_head(n = 800) %>%
  pull(peak_id)

heat_mat <- as.matrix(counts[top_signature, ])
heat_mat_z <- t(scale(t(heat_mat)))
heat_mat_z[is.na(heat_mat_z)] <- 0
ann_col <- sample_info %>%
  select(sample, condition, donor) %>%
  as.data.frame()
rownames(ann_col) <- ann_col$sample
ann_col$sample <- NULL

pheatmap(
  heat_mat_z,
  show_rownames = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  main = "Top IL-10-suppressed LPS-induced ATAC peaks"
)
```

The signature definition is deliberately stricter than "any negative IL-10
effect": the peak must first be LPS-induced, then decreased by IL-10 in the LPS
context. This avoids calling constitutively variable or IL-10-only peaks part of
the anti-inflammatory program.

Using a strong LPS-induced threshold, `r suppression_summary$lps_induced` peaks
open with LPS and `r suppression_summary$strong_lps_il10_suppressed` of those
are suppressed by IL-10. A broader but still donor-adjusted definition identifies
`r suppression_summary$broad_lps_il10_suppressed` LPS-induced peaks with IL-10
reversal. The heatmap shows that this is not driven by one donor; the signature
has a coherent condition-level pattern across the paired samples.

I do not require the same peak to be FDR-significant in both LPS and TNF
suppression contrasts. The README warns that this collapses the signature for
reasons of power and stimulus biology. A better scientific question is whether
the LPS-defined IL-10-suppressed peaks tend to move in the same direction in the
TNF setting. That is closer to the paper's conclusion: IL-10 repeatedly targets
IRF/interferon-linked responses, but the strength and detectability of the
signal differ by stimulus and assay.

## MCQ 4: IL-10 suppression signature

The LPS-defined IL-10 suppression signature contains thousands of LPS-induced
peaks that close with IL-10, and most of the strong signature moves in the same
direction in the TNF suppression contrast even though the signature is not
defined by requiring co-significance. Which conclusion best captures the
regulatory interpretation?

A. IL-10 targets a directionally consistent suppression program across stimuli,
but the analysis should derive a stable foreground from the stronger LPS arm and
use TNF as a sign-consistency test rather than a strict co-FDR filter.

B. The TNF arm should be ignored because TNF receptor signaling and TLR4
signaling cannot share downstream chromatin regulation.

C. Only peaks significant in both LPS and TNF suppression contrasts should be
used downstream; requiring co-significance is the only durable definition of
stimulus-independent IL-10 biology.

D. The largest IL-10 effect is best interpreted as a global reduction in
accessibility, because the signature heatmap shows lower signal in IL-10-treated
samples.

**Answer: A.** The result supports a real IL-10 suppression axis with asymmetric
detectability across stimuli. This framing preserves enough LPS-defined peaks
for motif and pathway analysis while still testing whether the TNF arm agrees in
direction.

# Peak-to-gene annotation

```{r import-gencode}
gtf_genes_rds <- file.path(ref_dir, "gencode.v45.annotation.genes.rds")
if (file.exists(gtf_genes_rds)) {
  genes_gr <- readRDS(gtf_genes_rds)
} else {
  genes_gr <- rtracklayer::import(gencode_gtf_gz, feature.type = "gene")
  saveRDS(genes_gr, gtf_genes_rds)
}

genes_tbl <- as_tibble(as.data.frame(genes_gr)) %>%
  transmute(
    gene_chrom = seqnames,
    gene_start = start,
    gene_end = end,
    strand = strand,
    gene_id = gene_id,
    gene_name = gene_name,
    gene_type = gene_type,
    tss = if_else(strand == "-", gene_end, gene_start)
  )

peak_gr <- GRanges(seqnames = bed$chrom, ranges = IRanges(start = bed$start + 1, end = bed$end))
names(peak_gr) <- bed$peak_id
gene_gr <- GRanges(seqnames = genes_tbl$gene_chrom, ranges = IRanges(genes_tbl$gene_start, genes_tbl$gene_end), strand = genes_tbl$strand)
names(gene_gr) <- genes_tbl$gene_name

nearest_idx <- nearest(peak_gr, gene_gr, ignore.strand = TRUE)
nearest_genes <- genes_tbl[nearest_idx, ]
peak_mid <- start(peak_gr) + floor((width(peak_gr) - 1) / 2)
distance_to_tss <- abs(peak_mid - nearest_genes$tss)

promoter_gr <- promoters(gene_gr, upstream = 2000, downstream = 500)
promoter_hits <- countOverlaps(peak_gr, promoter_gr, ignore.strand = TRUE) > 0
gene_body_hits <- countOverlaps(peak_gr, gene_gr, ignore.strand = TRUE) > 0

annotation_tbl <- bed %>%
  mutate(
    nearest_gene = nearest_genes$gene_name,
    nearest_gene_id = nearest_genes$gene_id,
    nearest_gene_type = nearest_genes$gene_type,
    distance_to_tss = distance_to_tss,
    regulatory_class = case_when(
      promoter_hits ~ "promoter_proximal",
      gene_body_hits ~ "gene_body",
      distance_to_tss <= 50000 ~ "distal_within_50kb",
      TRUE ~ "distal_gt_50kb"
    )
  ) %>%
  left_join(signature_tbl, by = c("peak_id", "chrom", "start", "end", "width"))

annotation_summary <- annotation_tbl %>%
  mutate(set = if_else(il10_lps_suppressed, "IL10_suppressed_signature", "Other_peaks")) %>%
  count(set, regulatory_class) %>%
  group_by(set) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()
annotation_summary
```

```{r annotation-plots}
ggplot(annotation_summary, aes(regulatory_class, fraction, fill = set)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = NULL,
    y = "Fraction of peaks",
    fill = NULL,
    title = "IL-10-suppressed peaks include promoter and distal regulatory contexts"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

genes_of_interest <- c("IFIT1", "IFIT2", "IFIT3", "ISG15", "CXCL10", "IRF1", "IRF5", "TNF", "IL6", "IL1B")
annotation_tbl %>%
  filter(il10_lps_suppressed, nearest_gene %in% genes_of_interest) %>%
  arrange(distance_to_tss) %>%
  select(peak_id, regulatory_class, nearest_gene, distance_to_tss, lps_logFC, il10_lps_logFC, il10_tnf_logFC) %>%
  head(25)
```

Nearest-gene annotation is useful but not sufficient. A promoter-proximal peak
near an interferon-stimulated gene is more interpretable than a distal peak that
happens to be nearest to a large gene desert neighbor; conversely, a distal
enhancer may regulate a gene that is not the nearest TSS. Because this bundle
does not include chromatin loops, enhancer-promoter maps, RNA-seq per donor, or
IRF1 CUT&RUN tracks, I treat the annotation as a prioritization layer rather
than proof of target-gene regulation.

The paper's broader mechanism helps calibrate the interpretation. IL-10 does not
need to erase all inflammatory chromatin to suppress inflammatory gene
expression. Peaks near ISG and IRF-linked loci are particularly relevant because
the full study tied IL-10 suppression to IRF1/IRF5 and the IFN-beta amplification
loop, while many NF-kB/AP-1-accessible loci were more preserved.

In this dataset, the IL-10-suppressed signature is not purely promoter-proximal:
`r scales::percent(annotation_summary$fraction[annotation_summary$set == "IL10_suppressed_signature" & annotation_summary$regulatory_class == "gene_body"], accuracy = 0.1)`
falls in gene bodies and `r scales::percent(sum(annotation_summary$fraction[annotation_summary$set == "IL10_suppressed_signature" & annotation_summary$regulatory_class %in% c("distal_within_50kb", "distal_gt_50kb")]), accuracy = 0.1)`
is distal by this simple annotation. That pattern fits enhancer-rich immune
regulation and makes proximity-based interpretation useful but incomplete. The
right biological readout is therefore not "nearest gene equals target gene"; it
is whether the annotated regulatory neighborhoods are coherent with the
IRF/interferon biology tested in the motif and pathway sections.

## MCQ 5: peak-to-gene annotation

The IL-10-suppressed peaks include promoter-proximal, gene-body and distal
regulatory contexts, with examples near interferon-stimulated and inflammatory
genes. Which annotation strategy best supports downstream biological
interpretation?

A. Treat nearest-gene labels as regulatory hypotheses and interpret them with
promoter overlap, gene-body overlap, distance to TSS, gene biotype and pathway
coherence, especially for distal immune enhancers.

B. Restrict the analysis to promoter-proximal peaks only, because distal ATAC
peaks cannot contribute to interferon or inflammatory gene regulation.

C. Use the nearest protein-coding gene as the confirmed target for every
suppressed peak, then interpret pathway enrichment as direct gene repression.

D. Prioritize the genes with the smallest TSS distance even if the motif and
pathway results point to a different regulatory program.

**Answer: A.** The annotation pattern supports enhancer-rich immune regulation.
Nearest-gene labels are valuable for enrichment and hypothesis generation, but
the biological claim becomes stronger only when proximity agrees with regulatory
class, motif grammar and pathway coherence.

# TF motif enrichment

```{r motif-background}
mean_lps_accessibility <- rowMeans(counts[, sample_info$condition == "LPS", drop = FALSE])
motif_tbl <- annotation_tbl %>%
  mutate(mean_lps_accessibility = mean_lps_accessibility[peak_id])

foreground_ids <- motif_tbl %>%
  filter(il10_lps_suppressed) %>%
  arrange(il10_lps_fdr, il10_lps_logFC) %>%
  pull(peak_id)

candidate_bg <- motif_tbl %>%
  filter(
    lps_induced,
    !il10_lps_broad_suppressed,
    !peak_id %in% foreground_ids
  ) %>%
  mutate(
    width_bin = ntile(width, 10),
    access_bin = ntile(mean_lps_accessibility, 10)
  )

fg_bins <- motif_tbl %>%
  filter(peak_id %in% foreground_ids) %>%
  mutate(width_bin = ntile(width, 10), access_bin = ntile(mean_lps_accessibility, 10)) %>%
  count(width_bin, access_bin, name = "fg_n")

background_ids <- candidate_bg %>%
  inner_join(fg_bins, by = c("width_bin", "access_bin")) %>%
  group_by(width_bin, access_bin) %>%
  mutate(random_order = runif(n())) %>%
  arrange(random_order, .by_group = TRUE) %>%
  filter(row_number() <= 2 * first(fg_n)) %>%
  ungroup() %>%
  pull(peak_id)

motif_sets <- tibble(
  set = c("foreground", "background"),
  n = c(length(foreground_ids), length(background_ids))
)
motif_sets
```

```{r motif-scan}
motif_cache <- file.path(ref_dir, "jaspar2024_il10_signature_motif_results.rds")
if (file.exists(motif_cache)) {
  motif_enrichment <- readRDS(motif_cache)
} else {
  all_motif_ids <- c(foreground_ids, background_ids)
  all_gr <- peak_gr[all_motif_ids]
  pwms <- TFBSTools::getMatrixSet(
    JASPAR2024::db(JASPAR2024::JASPAR2024()),
    opts = list(collection = "CORE", tax_group = "vertebrates", all_versions = FALSE)
  )
  motif_ix <- motifmatchr::matchMotifs(
    pwms,
    all_gr,
    genome = BSgenome.Hsapiens.UCSC.hg38,
    out = "matches",
    p.cutoff = 5e-5
  )
  motif_mat <- as.matrix(motifmatchr::motifMatches(motif_ix))
  is_fg <- rownames(motif_mat) %in% foreground_ids
  motif_enrichment <- map_dfr(seq_along(pwms), function(i) {
    present_fg <- sum(motif_mat[is_fg, i])
    absent_fg <- sum(!motif_mat[is_fg, i])
    present_bg <- sum(motif_mat[!is_fg, i])
    absent_bg <- sum(!motif_mat[!is_fg, i])
    ft <- fisher.test(matrix(c(present_fg, absent_fg, present_bg, absent_bg), nrow = 2, byrow = TRUE))
    tibble(
      motif_id = TFBSTools::ID(pwms[[i]]),
      motif_name = TFBSTools::name(pwms[[i]]),
      foreground_with_motif = present_fg,
      background_with_motif = present_bg,
      foreground_fraction = present_fg / (present_fg + absent_fg),
      background_fraction = present_bg / (present_bg + absent_bg),
      odds_ratio = unname(ft$estimate),
      p_value = ft$p.value
    )
  }) %>%
    mutate(fdr = p.adjust(p_value, method = "BH")) %>%
    arrange(fdr, desc(odds_ratio))
  saveRDS(motif_enrichment, motif_cache)
}

motif_enrichment %>%
  filter(odds_ratio > 1) %>%
  slice_head(n = 20)
```

```{r motif-plot}
top_motifs <- motif_enrichment %>%
  filter(odds_ratio > 1, is.finite(odds_ratio)) %>%
  slice_head(n = 15) %>%
  mutate(motif_label = paste0(motif_name, " (", motif_id, ")"))

ggplot(top_motifs, aes(reorder(motif_label, odds_ratio), odds_ratio, fill = -log10(fdr))) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Motif odds ratio",
    fill = "-log10 FDR",
    title = "Motifs enriched in IL-10-suppressed LPS-induced peaks"
  ) +
  theme_bw()
```

The motif analysis tests the regulatory grammar of the IL-10-sensitive peak set.
The foreground is the strong LPS-induced, IL-10-suppressed signature
(`r length(foreground_ids)` peaks). The background is not all merged peaks; it is
LPS-induced accessible chromatin that is not broadly IL-10-suppressed, sampled
with width and LPS-accessibility matching (`r length(background_ids)` peaks).
That background asks the sharper question for an R&D audience: among
inflammatory regulatory elements that open with LPS, what distinguishes the
subset that IL-10 can close?

The strongest enrichments are IRF/ISRE-family and STAT1/STAT2-like motifs.
For example, `r motif_enrichment$motif_name[1]` is present in
`r scales::percent(motif_enrichment$foreground_fraction[1], accuracy = 0.1)` of
foreground peaks versus
`r scales::percent(motif_enrichment$background_fraction[1], accuracy = 0.1)` of
matched background peaks, with an odds ratio of
`r round(motif_enrichment$odds_ratio[1], 2)` after BH correction across tested
motifs. This aligns strongly with Mishra et al.: IL-10 does not simply remove
all inflammatory enhancer grammar; it preferentially marks the IRF/interferon
arm of the response as suppressible.

The careful limitation is TF identity. A JASPAR IRF motif does not prove IRF1
binding, IRF5 binding, or displacement of a specific IRF at that site. The paper
could make that stronger claim because it added CUT&RUN, footprinting and
perturbation. From this ATAC matrix, the defensible conclusion is that the
IL-10-sensitive chromatin compartment is enriched for IRF/ISRE-compatible
sequence grammar, not that a single TF has been causally assigned.

## MCQ 6: TF motif enrichment

The motif foreground is the LPS-induced IL-10-suppressed signature, and the
matched background is LPS-induced chromatin that remains accessible with IL-10.
The top enriched motifs are IRF/ISRE-family and STAT1/STAT2-like motifs. Which
interpretation best fits these results?

A. IL-10-sensitive peaks are distinguished from other LPS-induced accessible
regions by IRF/interferon-compatible sequence grammar, supporting the paper's
model that IL-10 preferentially suppresses the IRF/IFN regulatory arm rather
than uniformly erasing inflammatory accessibility.

B. IRF motif enrichment proves that IRF1 is directly bound at every foreground
peak and that IRF5 is not involved.

C. Because the foreground and background are both LPS-induced, any IRF
enrichment must be a technical artifact of peak width rather than biology.

D. The absence of a uniquely dominant NF-kB motif means LPS did not activate
canonical inflammatory chromatin in this dataset.

**Answer: A.** The matched background makes the result more specific: this is
not merely "inflammatory peaks contain motifs," but that the IL-10-sensitive
fraction of the LPS response is especially IRF/ISRE-like. The result supports
the paper's regulatory model without overclaiming exact TF occupancy from motif
sequence alone.

# Pathway enrichment and regulatory interpretation

```{r hallmark-parse}
parse_gmt <- function(path) {
  lines <- readLines(path)
  pieces <- strsplit(lines, "\t")
  tibble(
    pathway = vapply(pieces, `[[`, character(1), 1),
    genes = lapply(pieces, function(x) x[-c(1, 2)])
  )
}

hallmark <- parse_gmt(hallmark_gmt)
signature_genes <- annotation_tbl %>%
  filter(il10_lps_suppressed, nearest_gene_type == "protein_coding") %>%
  pull(nearest_gene) %>%
  unique()

background_genes <- annotation_tbl %>%
  filter(lps_induced, nearest_gene_type == "protein_coding") %>%
  pull(nearest_gene) %>%
  unique()

enrich_hyper <- function(path_genes, query, universe) {
  path_genes <- intersect(path_genes, universe)
  query <- intersect(query, universe)
  a <- length(intersect(path_genes, query))
  b <- length(query) - a
  c <- length(path_genes) - a
  d <- length(universe) - a - b - c
  if (a == 0 || length(path_genes) < 5) {
    return(tibble(overlap = a, pathway_size = length(path_genes), p_value = 1, odds_ratio = NA_real_))
  }
  ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2, byrow = TRUE), alternative = "greater")
  tibble(overlap = a, pathway_size = length(path_genes), p_value = ft$p.value, odds_ratio = unname(ft$estimate))
}

hallmark_enrichment <- hallmark %>%
  mutate(stats = map(genes, enrich_hyper, query = signature_genes, universe = background_genes)) %>%
  select(pathway, stats) %>%
  unnest(stats) %>%
  mutate(fdr = p.adjust(p_value, method = "BH")) %>%
  arrange(fdr, desc(odds_ratio))

hallmark_enrichment %>% slice_head(n = 20)
```

```{r hallmark-plot}
top_pathways <- hallmark_enrichment %>%
  filter(overlap > 0, is.finite(odds_ratio)) %>%
  slice_head(n = 12) %>%
  mutate(pathway = gsub("^HALLMARK_", "", pathway))

ggplot(top_pathways, aes(reorder(pathway, odds_ratio), odds_ratio, fill = -log10(fdr))) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Odds ratio",
    fill = "-log10 FDR",
    title = "Hallmark enrichment among genes near IL-10-suppressed peaks"
  ) +
  theme_bw()
```

The pathway result closes the loop between chromatin statistics and monocyte
biology. Genes near IL-10-suppressed peaks are most strongly enriched for
Hallmark interferon gamma response and interferon alpha response
(FDR `r signif(hallmark_enrichment$fdr[hallmark_enrichment$pathway == "HALLMARK_INTERFERON_GAMMA_RESPONSE"], 3)`
and `r signif(hallmark_enrichment$fdr[hallmark_enrichment$pathway == "HALLMARK_INTERFERON_ALPHA_RESPONSE"], 3)`,
respectively). TNF/NF-kB signaling is present but weaker
(FDR `r signif(hallmark_enrichment$fdr[hallmark_enrichment$pathway == "HALLMARK_TNFA_SIGNALING_VIA_NFKB"], 3)`).
That ranking matters: it mirrors the paper's central observation that IL-10 more
strongly suppresses IFN/ISG biology while inflammatory NF-kB programs are
modulated but not simply eliminated.

The integrated model is therefore selective rather than global. IL-10
preferentially suppresses LPS-induced accessible elements with IRF/ISRE-like
motif grammar, and those elements annotate near genes in interferon and
inflammatory response pathways. In the full Mishra et al. study, that chromatin
pattern is connected to suppression of IRF1/IRF5 and the IFN-beta amplification
loop. In this take-home matrix, I can recapitulate the regulatory footprint of
that model, but I cannot independently prove TF occupancy, H3K27ac loss, IFNB
production, or enhancer-promoter looping.

A practical follow-up would be to layer RNA-seq or perturbation data onto this
signature: prioritize suppressed ATAC peaks near ISG loci, test whether linked
genes lose expression with IL-10, and use IRF1/IRF5 or STAT1/STAT2 occupancy data
to separate motif-family inference from TF-specific mechanism. That is the
decision point where this report moves from regulatory interpretation to causal
model testing.

## MCQ 7: pathway and regulatory interpretation

The IL-10-suppressed ATAC signature is enriched for IRF/ISRE-like motifs and
maps near genes enriched most strongly for interferon alpha/gamma response, with
weaker but present TNF/NF-kB pathway signal. Which final model best integrates
these findings with Mishra et al.?

A. IL-10 globally closes monocyte chromatin; interferon pathways appear enriched
only because ISG loci are generally more accessible than inflammatory loci.

B. IL-10 preferentially suppresses an IRF/interferon-linked component of the
LPS-induced chromatin response, with secondary inflammatory pathway involvement;
this is consistent with the paper's IRF1/IRF5 and IFN-beta amplification-loop
model, while remaining short of proving TF occupancy from ATAC alone.

C. IL-10 blocks canonical NF-kB activation genome-wide, and the interferon motif
signal is downstream of complete NF-kB shutdown.

D. The pathway result should be ignored because nearest-gene annotation is
imperfect; motif enrichment alone is sufficient to define the mechanism.

**Answer: B.** The chromatin, motif and pathway layers converge on selective
IRF/interferon suppression with inflammatory pathway involvement. The strongest
claim from this report is mechanistic alignment with Mishra et al.; causal TF
assignment still requires the orthogonal data types used in the paper.

# Reproducibility

```{r session-info}
sessionInfo()
```
